---
title: "Exercise 3: Trends, Seasons, Cycles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br>
*This document serves as a presentation of exercise 3 for course Introduction to Forecasting Models.
The code provided in this document matches the source code of the solution itself.*

<br>

### Setup environment

As the first step it is a good idea to clear and setup the environment.

1. Delete all variables
2. Reset the GUI for plotting
3. Set the path to match the location of the source code
<br>
```{r, results='hide'}
rm(list = ls())
dev.off()
setwd('C:/Users/Adam Cihlář/Desktop/materiály/ForecastingModels/assignments/exercise3')
```

4. Load all necessary packages

```{r, results='hide', warning=FALSE, message=FALSE}
library(tidyverse)
library(readxl)
library(mltools)
library(data.table)
library(gridExtra)
library(stargazer)
```


```{r}
data <- read_xls(path = '../data/Sales new cars US.xls', skip = 10) %>%
    mutate(observation_date = as.Date(observation_date),
           m = lubridate::month(observation_date),
           time_index = c(1:nrow(.)),
           time_index_2 = time_index^2,
           y = TOTALNSA,
           y_1 = dplyr::lag(y, n = 1),
           y_2 = dplyr::lag(y, n = 2),
           y_3 = dplyr::lag(y, n = 3),
           y_4 = dplyr::lag(y, n = 4),
           y_12 = dplyr::lag(y, n = 12),
           y_13 = dplyr::lag(y, n = 13),
           )

data$m <- as.factor(data$m)
data <- as_tibble(one_hot(as.data.table(data)))

head(data)
```